Multi-temporal drought and its effects on land cover change and vegetation productivity in continental Chile

### 2024-03-13 article/paper_drought_Chile2023.qmd
first draft

title: Drought, vegetation productivity, and land cover change in Chile

@@

@@

  Central Chile has been the focus of research studies due to the persistent decrease in water supply, which is impacting the hydrological system and vegetation development. This persistent period of water scarcity has been defined as a megadrought. Our objective is to assess the impact of drought on LULCC (land use land cover change) over continental Chile using drought indices of water supply and demand, soil moisture, and their impact on vegetation productivity. The monthly ERA5-Land (ERA5L) variables for precipitation, temperature, and soil moisture were used. From 2001 to 2022, we used the land cover MODIS product MCD12Q1, and from 2000 to 2023, we used the NDVI (Normalized Difference Vegetation Index) product MOD13A3 collection 6.1. As drought indices, we compute the standardized anomaly of cumulative NDVI (zcNDVI), the Standardized Precipitation Evapotranspiration Index (SPEI), the Evaporative Demand Drought Index (EDDI), and the Standardized Soil Moisture Index (SSI). These indices were calculated for time scales of 1, 3, 6, 12, 24, and 36 months, except for zcNDVI, which was for 6 months. We analyze the trend for LULCC, vegetation productivity, and drought indices. Also, we analyzed the temporal correlation of SPI, SPEI, EDDI, and SSI with zcNDVI to gain insights into the impact of water supply and demand on vegetation productivity. Our results showed that LULCC was highest in "Centro," "Sur," and "Austral,” with 36%, 31%, and 34% of change in the surface type, respectively. The EDDI shows that water demand has increased for all zones, with a major increase in "Norte Grande." The drought indices of water supply and soil moisture evidence a decreasing trend, which decreases at longer time scales, from "Norte Grande" to "Sur." "Austral" is the only zone that shows an increase in supply. Vegetation productivity measures by zcNDVI present a negative trend in "Norte Chico" and "Centro." On the other hand, forests seem to be the most resistant to drought. The types that show to be most affected by variation in climate conditions are shrublands, savannas, and croplands. The drought indices that have the capability of explaining to a major degree the variance in vegetation productivity are SSI-12, followed by SPEI-24 and SPEI-12 in "Norte Chico" and "Centro." The results indicate that “Norte Chico” and “Zona Central” are the most sensitive regions to water supply deficits lasting longer than a year. Our results can help develop a robust vegetation productivity forecasting model for land cover classes in Chile.

@@

## Trend of drought and land cover change

@@

To derive the drought indices, first we must calculate the sum of the variables with regard to the time scale (s). In this case, for generalization purposes, we will use $V$, referring to variables $P$, $AED$, $D$, $NDVI$, and $SM$ (#ref{tab-desEOD}). We cumulated each $V$  over the time series of $n$ values (months), and for the time scales $s$:

@@

@@

For vegetation, we obtained and use hereafter five macroclasses of land cover from IGBP MODIS: forest, shrubland, savanna, grassland, and croplands. [@fig-studyArea]c shows the spatial distribution of the macroclasses through Chile for the year 2022. [@fig-studyArea]d shows the macroclasses of land cover persistance (80%) during 2021–2022, respectively (#ref{tab-landcoverSurf}). Within continental Chile, barren land is the land cover class with the highest surface area (277,870 km2). The largest type of vegetation, with 137,085 km2, is forest. Grassland (74,247 km2), savanna (55,206 km2), shrubland (25,341 km2), and cropland (3,146 km2) are the other types (Table 2). The macrozones with major changes for 2001–2022 were "Centro," "Sur," and "Austral," with 36%, 31%, and 34% of their surface changing the type of land cover, respectively (@fig-studyArea and #ref{tab-landcoverTrend}). @fig-LCprop shows the summary of the proportion of surface per land cover class and macrozone, derived from the persistance mask over continental Chile.

@@

The "Norte Chico" shows an increase in barren land of 111 $km^2 yr^{-1}$  and a reduction in the class savanna of 70 $km^2 yr^{-1}$. In the "Centro" and "Sur," there are changes with an important reduction in savanna (136 to 318 $km^2 yr^{-1}$ ), and an increase in shrubland and grassland. Showing a change for more dense vegetation types. It appears to be a shift in the area cultivated (croplands) from the "Centro" to the "Sur." Also, there is a high increase in forest (397 $km^2 yr^{-1}$ ) in the "Sur," replacing the savanna lost (#ref{tab-landcoverTrend}).

## Trend on drought and its relation to land cover change

@@

In #ref{tab-corlandcover}, we aggregate per macrozone and landcover the correlation analysis presented in @fig-corTimeScale and @fig-corPerson. According to what is shown, forests seem to be the most resistant to drought. Showing that only "Centro" is slightly (rsq = 0.25) impacted by a 12-month soil moisture deficit (SSI-12). In the "Norte Chico" and to a lesser extent in the "Norte Grande," it is evident that a SSI-12 with a rsq = 0.45 and a decrease in water supply (SPI-36 and SPEI-24 with rsq = 0.28 and 0.34, respectively) have an impact on grasslands. However, this type was unaffected by soil moisture, water supply, or demand in macrozones further south. The types that show to be most affected by variation in climate conditions are shrublands, savannas, and croplands. For savannas in "Norte Chico," the SSI-12 and SPI-24 reached an rsq of 0.74 and 0.58, respectively. This value decreases to the south, but the SSI-12 is still the variable explaining more of the variation in vegetation productivity (rsq = 0.45 in "Centro" and 0.2 in "Sur"). In the case of croplands, the SPEI-12, SPI-36, and SSI-12 explain between 45% and 66% of the variability in "Norte Chico." The type of land most impacted by climatic variation was shrubland, where soil moisture explained 59% and precipitation, 37%, in "Norte Chico" and "Centro," with SSI-12 being the most relevant variable, then SPI-36 in "Norte Chico" and SPI-24 in "Sur."

>>>>

title: Multi-temporal drought and its interaction with land cover change and vegetation productivity in continental Chile

@@

    email: francisco.fernandez@uss.cl

@@

 The north-central region of Chile has been the focus of research studies due to the persistent decrease in water supply, which is impacting the hydrological system and vegetation development. This persistent period of water scarcity has been defined as a megadrought. The aim of our study is to evaluate the land cover change over continental Chile and to examine how this is connected to drought indices of water supply, atmospheric evaporative demand (AED), soil moisture, and their effects on vegetation productivity. The drought indices were derived using monthly ERA5-Land reanalysis data spanning from 1981 to 2023. The Moderate-Resolution Imaging Spectroradiometer (MODIS) datasets were utilized to obtain information on annual land cover and monthly vegetation productivity. We analyzed short- (1, 3, 6 months) to long-term (12, 24, 36 months) time scales of drought. Our results showed that land cover change was highest in the south-central part of the country, reaching changes as high as 36% in the surface type. The water demand has increased for the whole country, with a major increase in the north. The AED and soil moisture evidence a decreasing trend, which decreases at longer time scales and from north to south. The extreme south part of the country shows an increase in supply. Vegetation productivity has a negative trend in the north-central region for all land cover types. On the other hand, forests seem to be the most resistant type to drought. The types that show to be most affected by variation in climate conditions are shrublands, savannas, and croplands. The drought indices that have the capability of explaining to a major degree the variance in vegetation productivity are the ones that consider soil moisture for twelve months and the combined effect of precipitation and AED for 24 and 12 months. The results indicate that the north-central region is the most sensitive to water supply deficits lasting longer than a year. 

@@

## Trend of drought and interaction to land cover change

@@

To derive the drought indices, first we must calculate the sum of the variables with regard to the time scale (s). In this case, for generalization purposes, we will use $V$, referring to variables $P$, $AED$, $D$, $NDVI$, and $SM$ (Table \ref{tab-desEOD}). We cumulated each $V$  over the time series of $n$ values (months), and for the time scales $s$:

@@

### Intercation drought indices and land cover change

We wanted to explore the relationship between the trend in land cover classes and the trend in the drought indices. Regarding the trends of land cover obtained (see Section 3.4.5), we divide each trend (km2) per land cover class and macrozone by the total surface. That way, we relativize the trend regarding the total land cover surface type per zone. To the trends of the drought indices over Chile, we applied the mask of persistent land cover macroclass and aggregated the trend for land cover class and macrozone. We then explore visually the relationship between the trend of drought indices for the short- to long-term and the trend of land cover.

@@

For vegetation, we obtained and use hereafter five macroclasses of land cover from IGBP MODIS: forest, shrubland, savanna, grassland, and croplands. [@fig-studyArea]c shows the spatial distribution of the macroclasses through Chile for the year 2022. [@fig-studyArea]d shows the macroclasses of land cover persistance (80%) during 2021–2022, respectively (Table \ref{tab-landcoverSurf}). Within continental Chile, barren land is the land cover class with the highest surface area (277,870 km2). The largest type of vegetation, with 137,085 km2, is forest. Grassland (74,247 km2), savanna (55,206 km2), shrubland (25,341 km2), and cropland (3,146 km2) are the other types (Table \ref{tab-landcoverSurf}). The macrozones with major changes for 2001–2022 were "Centro," "Sur," and "Austral," with 36%, 31%, and 34% of their surface changing the type of land cover, respectively (@fig-studyArea and Table \ref{tab-landcoverTrend}). @fig-LCprop shows the summary of the proportion of surface per land cover class and macrozone, derived from the persistance mask over continental Chile.

@@

The "Norte Chico" shows an increase in barren land of 111 $km^2 yr^{-1}$  and a reduction in the class savanna of 70 $km^2 yr^{-1}$. In the "Centro" and "Sur," there are changes with an important reduction in savanna (136 to 318 $km^2 yr^{-1}$ ), and an increase in shrubland and grassland. Showing a change for more dense vegetation types. It appears to be a shift in the area cultivated (croplands) from the "Centro" to the "Sur." Also, there is a high increase in forest (397 $km^2 yr^{-1}$ ) in the "Sur," replacing the savanna lost (Table \ref{tab-landcoverTrend}).

## Trend of drought and interaction to land cover change

@@

In Table \ref{tab-corlandcover}, we aggregate per macrozone and landcover the correlation analysis presented in @fig-corTimeScale and @fig-corPerson. According to what is shown, forests seem to be the most resistant to drought. Showing that only "Centro" is slightly (rsq = 0.25) impacted by a 12-month soil moisture deficit (SSI-12). In the "Norte Chico" and to a lesser extent in the "Norte Grande," it is evident that a SSI-12 with a rsq = 0.45 and a decrease in water supply (SPI-36 and SPEI-24 with rsq = 0.28 and 0.34, respectively) have an impact on grasslands. However, this type was unaffected by soil moisture, water supply, or demand in macrozones further south. The types that show to be most affected by variation in climate conditions are shrublands, savannas, and croplands. For savannas in "Norte Chico," the SSI-12 and SPI-24 reached an rsq of 0.74 and 0.58, respectively. This value decreases to the south, but the SSI-12 is still the variable explaining more of the variation in vegetation productivity (rsq = 0.45 in "Centro" and 0.2 in "Sur"). In the case of croplands, the SPEI-12, SPI-36, and SSI-12 explain between 45% and 66% of the variability in "Norte Chico." The type of land most impacted by climatic variation was shrubland, where soil moisture explained 59% and precipitation, 37%, in "Norte Chico" and "Centro," with SSI-12 being the most relevant variable, then SPI-36 in "Norte Chico" and SPI-24 in "Sur."

### 2024-03-13 article/paper_drought_Chile2023.tex
first draft

\usepackage{lineno}\linenumbers \usepackage{multirow} \usepackage{lscape} \newcommand{\blandscape}{\begin{landscape}} \newcommand{\elandscape}{\end{landscape}}
\makeatletter
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[skins,breakable]{tcolorbox}}
\makeatother
\makeatletter
\@ifundefined{shadecolor}{\definecolor{shadecolor}{rgb}{.97, .97, .97}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\makeatother
\journal{Journal Name}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage[]{natbib}
\bibliographystyle{elsarticle-harv}
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Drought, vegetation productivity, and land cover change in Chile},
  pdfauthor={Francisco Zambrano; Anton Vrieling; Francisco Fernández; Francisco Meza; Alejandro Venegas-González; Iongel Duran-Llacer; Nicolas Raab; Dylan Craven},
  pdfkeywords={drought, land cover change, satellite},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

\begin{frontmatter}
\title{Drought, vegetation productivity, and land cover change in Chile}
\author[1]{Francisco Zambrano%
\corref{cor1}%
\fnref{fn1}}
 \ead{francisco.zambrano@umayor.cl} 
\author[2]{Anton Vrieling%
%
}
 \ead{a.vrieling@utwente.nl} 
\author[3]{Francisco Fernández%
%
}
 \ead{a.vrieling@utwente.nl} 
\author[]{Francisco Meza%
%
}

@@

\begin{abstract}
Central Chile has been the focus of research studies due to the
persistent decrease in water supply, which is impacting the hydrological
system and vegetation development. This persistent period of water
scarcity has been defined as a megadrought. Our objective is to assess
the impact of drought on LULCC (land use land cover change) over
continental Chile using drought indices of water supply and demand, soil
moisture, and their impact on vegetation productivity. The monthly
ERA5-Land (ERA5L) variables for precipitation, temperature, and soil
moisture were used. From 2001 to 2022, we used the land cover MODIS
product MCD12Q1, and from 2000 to 2023, we used the NDVI (Normalized
Difference Vegetation Index) product MOD13A3 collection 6.1. As drought
indices, we compute the standardized anomaly of cumulative NDVI
(zcNDVI), the Standardized Precipitation Evapotranspiration Index
(SPEI), the Evaporative Demand Drought Index (EDDI), and the
Standardized Soil Moisture Index (SSI). These indices were calculated
for time scales of 1, 3, 6, 12, 24, and 36 months, except for zcNDVI,
which was for 6 months. We analyze the trend for LULCC, vegetation
productivity, and drought indices. Also, we analyzed the temporal
correlation of SPI, SPEI, EDDI, and SSI with zcNDVI to gain insights
into the impact of water supply and demand on vegetation productivity.
Our results showed that LULCC was highest in ``Centro,'' ``Sur,'' and
``Austral,'' with 36\%, 31\%, and 34\% of change in the surface type,
respectively. The EDDI shows that water demand has increased for all
zones, with a major increase in ``Norte Grande.'' The drought indices of
water supply and soil moisture evidence a decreasing trend, which
decreases at longer time scales, from ``Norte Grande'' to ``Sur.''
``Austral'' is the only zone that shows an increase in supply.
Vegetation productivity measures by zcNDVI present a negative trend in
``Norte Chico'' and ``Centro.'' On the other hand, forests seem to be
the most resistant to drought. The types that show to be most affected
by variation in climate conditions are shrublands, savannas, and
croplands. The drought indices that have the capability of explaining to
a major degree the variance in vegetation productivity are SSI-12,
followed by SPEI-24 and SPEI-12 in ``Norte Chico'' and ``Centro.'' The
results indicate that ``Norte Chico'' and ``Zona Central'' are the most
sensitive regions to water supply deficits lasting longer than a year.
Our results can help develop a robust vegetation productivity
forecasting model for land cover classes in Chile.
\end{abstract}

@@

\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[borderline west={3pt}{0pt}{shadecolor}, frame hidden, boxrule=0pt, breakable, interior hidden, sharp corners, enhanced]}{\end{tcolorbox}}\fi

@@

\hypertarget{trend-of-drought-and-land-cover-change}{%
\subsection{Trend of drought and land cover
change}\label{trend-of-drought-and-land-cover-change}}

@@

To derive the drought indices, first we must calculate the sum of the
variables with regard to the time scale (s). In this case, for
generalization purposes, we will use \(V\), referring to variables
\(P\), \(AED\), \(D\), \(NDVI\), and \(SM\) (\#ref\{tab-desEOD\}). We
cumulated each \(V\) over the time series of \(n\) values (months), and
for the time scales \(s\):

@@

@@

For vegetation, we obtained and use hereafter five macroclasses of land
cover from IGBP MODIS: forest, shrubland, savanna, grassland, and
croplands. Figure~\ref{fig-studyArea}c shows the spatial distribution of
the macroclasses through Chile for the year 2022.
Figure~\ref{fig-studyArea}d shows the macroclasses of land cover
persistance (80\%) during 2021--2022, respectively
(\#ref\{tab-landcoverSurf\}). Within continental Chile, barren land is
the land cover class with the highest surface area (277,870 km2). The
largest type of vegetation, with 137,085 km2, is forest. Grassland
(74,247 km2), savanna (55,206 km2), shrubland (25,341 km2), and cropland
(3,146 km2) are the other types (Table 2). The macrozones with major
changes for 2001--2022 were ``Centro,'' ``Sur,'' and ``Austral,'' with
36\%, 31\%, and 34\% of their surface changing the type of land cover,
respectively (Figure~\ref{fig-studyArea} and
\#ref\{tab-landcoverTrend\}). Figure~\ref{fig-LCprop} shows the summary
of the proportion of surface per land cover class and macrozone, derived
from the persistance mask over continental Chile.

@@

The ``Norte Chico'' shows an increase in barren land of 111
\(km^2 yr^{-1}\) and a reduction in the class savanna of 70
\(km^2 yr^{-1}\). In the ``Centro'' and ``Sur,'' there are changes with
an important reduction in savanna (136 to 318 \(km^2 yr^{-1}\) ), and an
increase in shrubland and grassland. Showing a change for more dense
vegetation types. It appears to be a shift in the area cultivated
(croplands) from the ``Centro'' to the ``Sur.'' Also, there is a high
increase in forest (397 \(km^2 yr^{-1}\) ) in the ``Sur,'' replacing the
savanna lost (\#ref\{tab-landcoverTrend\}).

\hypertarget{trend-on-drought-and-its-relation-to-land-cover-change}{%
\subsection{Trend on drought and its relation to land cover
change}\label{trend-on-drought-and-its-relation-to-land-cover-change}}

@@

In \#ref\{tab-corlandcover\}, we aggregate per macrozone and landcover
the correlation analysis presented in Figure~\ref{fig-corTimeScale} and
Figure~\ref{fig-corPerson}. According to what is shown, forests seem to
be the most resistant to drought. Showing that only ``Centro'' is
slightly (rsq = 0.25) impacted by a 12-month soil moisture deficit
(SSI-12). In the ``Norte Chico'' and to a lesser extent in the ``Norte
Grande,'' it is evident that a SSI-12 with a rsq = 0.45 and a decrease
in water supply (SPI-36 and SPEI-24 with rsq = 0.28 and 0.34,
respectively) have an impact on grasslands. However, this type was
unaffected by soil moisture, water supply, or demand in macrozones
further south. The types that show to be most affected by variation in
climate conditions are shrublands, savannas, and croplands. For savannas
in ``Norte Chico,'' the SSI-12 and SPI-24 reached an rsq of 0.74 and
0.58, respectively. This value decreases to the south, but the SSI-12 is
still the variable explaining more of the variation in vegetation
productivity (rsq = 0.45 in ``Centro'' and 0.2 in ``Sur''). In the case
of croplands, the SPEI-12, SPI-36, and SSI-12 explain between 45\% and
66\% of the variability in ``Norte Chico.'' The type of land most
impacted by climatic variation was shrubland, where soil moisture
explained 59\% and precipitation, 37\%, in ``Norte Chico'' and
``Centro,'' with SSI-12 being the most relevant variable, then SPI-36 in
``Norte Chico'' and SPI-24 in ``Sur.''

>>>>

\usepackage{lineno}\linenumbers \usepackage{multirow} \usepackage{lscape} \newcommand{\blandscape}{\begin{landscape}} \newcommand{\elandscape}{\end{landscape}}
\makeatletter
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[skins,breakable]{tcolorbox}}
\makeatother
\makeatletter
\@ifundefined{shadecolor}{\definecolor{shadecolor}{rgb}{.97, .97, .97}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\makeatother
\journal{Journal Name}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage[]{natbib}
\bibliographystyle{elsarticle-harv}
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Multi-temporal drought and its effects on land cover change and vegetation productivity in continental Chile},
  pdfauthor={Francisco Zambrano; Anton Vrieling; Francisco Fernández; Francisco Meza; Iongel Duran-Llacer; Alejandro Venegas-González; Nicolas Raab; Dylan Craven},
  pdfkeywords={drought, land cover change, satellite},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

\begin{frontmatter}
\title{Multi-temporal drought and its effects on land cover change and
vegetation productivity in continental Chile}
\author[1]{Francisco Zambrano%
\corref{cor1}%
\fnref{fn1}}
 \ead{francisco.zambrano@umayor.cl} 
\author[2]{Anton Vrieling%
%
}
 \ead{a.vrieling@utwente.nl} 
\author[3]{Francisco Fernández%
%
}
 \ead{francisco.fernandez@uss.cl} 
\author[]{Francisco Meza%
%
}

@@

\begin{abstract}
The north-central region of Chile has been the focus of research studies
due to the persistent decrease in water supply, which is impacting the
hydrological system and vegetation development. This persistent period
of water scarcity has been defined as a megadrought. The aim of our
study is to evaluate the land cover change over continental Chile and to
examine how this is connected to drought indices of water supply,
atmospheric evaporative demand (AED), soil moisture, and their effects
on vegetation productivity. The drought indices were derived using
monthly ERA5-Land reanalysis data spanning from 1981 to 2023. The
Moderate-Resolution Imaging Spectroradiometer (MODIS) datasets were
utilized to obtain information on annual land cover and monthly
vegetation productivity. We analyzed short- (1, 3, 6 months) to
long-term (12, 24, 36 months) time scales of drought. Our results showed
that land cover change was highest in the south-central part of the
country, reaching changes as high as 36\% in the surface type. The water
demand has increased for the whole country, with a major increase in the
north. The AED and soil moisture evidence a decreasing trend, which
decreases at longer time scales and from north to south. The extreme
south part of the country shows an increase in supply. Vegetation
productivity has a negative trend in the north-central region for all
land cover types. On the other hand, forests seem to be the most
resistant type to drought. The types that show to be most affected by
variation in climate conditions are shrublands, savannas, and croplands.
The drought indices that have the capability of explaining to a major
degree the variance in vegetation productivity are the ones that
consider soil moisture for twelve months and the combined effect of
precipitation and AED for 24 and 12 months. The results indicate that
the north-central region is the most sensitive to water supply deficits
lasting longer than a year.
\end{abstract}

@@

\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, enhanced, interior hidden, borderline west={3pt}{0pt}{shadecolor}, sharp corners, breakable, frame hidden]}{\end{tcolorbox}}\fi

@@

\hypertarget{trend-of-drought-and-interaction-to-land-cover-change}{%
\subsection{Trend of drought and interaction to land cover
change}\label{trend-of-drought-and-interaction-to-land-cover-change}}

@@

To derive the drought indices, first we must calculate the sum of the
variables with regard to the time scale (s). In this case, for
generalization purposes, we will use \(V\), referring to variables
\(P\), \(AED\), \(D\), \(NDVI\), and \(SM\) (Table \ref{tab-desEOD}). We
cumulated each \(V\) over the time series of \(n\) values (months), and
for the time scales \(s\):

@@

\hypertarget{intercation-drought-indices-and-land-cover-change}{%
\subsubsection{Intercation drought indices and land cover
change}\label{intercation-drought-indices-and-land-cover-change}}

We wanted to explore the relationship between the trend in land cover
classes and the trend in the drought indices. Regarding the trends of
land cover obtained (see Section 3.4.5), we divide each trend (km2) per
land cover class and macrozone by the total surface. That way, we
relativize the trend regarding the total land cover surface type per
zone. To the trends of the drought indices over Chile, we applied the
mask of persistent land cover macroclass and aggregated the trend for
land cover class and macrozone. We then explore visually the
relationship between the trend of drought indices for the short- to
long-term and the trend of land cover.

@@

For vegetation, we obtained and use hereafter five macroclasses of land
cover from IGBP MODIS: forest, shrubland, savanna, grassland, and
croplands. Figure~\ref{fig-studyArea}c shows the spatial distribution of
the macroclasses through Chile for the year 2022.
Figure~\ref{fig-studyArea}d shows the macroclasses of land cover
persistance (80\%) during 2021--2022, respectively (Table
\ref{tab-landcoverSurf}). Within continental Chile, barren land is the
land cover class with the highest surface area (277,870 km2). The
largest type of vegetation, with 137,085 km2, is forest. Grassland
(74,247 km2), savanna (55,206 km2), shrubland (25,341 km2), and cropland
(3,146 km2) are the other types (Table \ref{tab-landcoverSurf}). The
macrozones with major changes for 2001--2022 were ``Centro,'' ``Sur,''
and ``Austral,'' with 36\%, 31\%, and 34\% of their surface changing the
type of land cover, respectively (Figure~\ref{fig-studyArea} and Table
\ref{tab-landcoverTrend}). Figure~\ref{fig-LCprop} shows the summary of
the proportion of surface per land cover class and macrozone, derived
from the persistance mask over continental Chile.

@@

The ``Norte Chico'' shows an increase in barren land of 111
\(km^2 yr^{-1}\) and a reduction in the class savanna of 70
\(km^2 yr^{-1}\). In the ``Centro'' and ``Sur,'' there are changes with
an important reduction in savanna (136 to 318 \(km^2 yr^{-1}\) ), and an
increase in shrubland and grassland. Showing a change for more dense
vegetation types. It appears to be a shift in the area cultivated
(croplands) from the ``Centro'' to the ``Sur.'' Also, there is a high
increase in forest (397 \(km^2 yr^{-1}\) ) in the ``Sur,'' replacing the
savanna lost (Table \ref{tab-landcoverTrend}).

\hypertarget{trend-of-drought-and-interaction-to-land-cover-change-1}{%
\subsection{Trend of drought and interaction to land cover
change}\label{trend-of-drought-and-interaction-to-land-cover-change-1}}

@@

In Table \ref{tab-corlandcover}, we aggregate per macrozone and
landcover the correlation analysis presented in
Figure~\ref{fig-corTimeScale} and Figure~\ref{fig-corPerson}. According
to what is shown, forests seem to be the most resistant to drought.
Showing that only ``Centro'' is slightly (rsq = 0.25) impacted by a
12-month soil moisture deficit (SSI-12). In the ``Norte Chico'' and to a
lesser extent in the ``Norte Grande,'' it is evident that a SSI-12 with
a rsq = 0.45 and a decrease in water supply (SPI-36 and SPEI-24 with rsq
= 0.28 and 0.34, respectively) have an impact on grasslands. However,
this type was unaffected by soil moisture, water supply, or demand in
macrozones further south. The types that show to be most affected by
variation in climate conditions are shrublands, savannas, and croplands.
For savannas in ``Norte Chico,'' the SSI-12 and SPI-24 reached an rsq of
0.74 and 0.58, respectively. This value decreases to the south, but the
SSI-12 is still the variable explaining more of the variation in
vegetation productivity (rsq = 0.45 in ``Centro'' and 0.2 in ``Sur'').
In the case of croplands, the SPEI-12, SPI-36, and SSI-12 explain
between 45\% and 66\% of the variability in ``Norte Chico.'' The type of
land most impacted by climatic variation was shrubland, where soil
moisture explained 59\% and precipitation, 37\%, in ``Norte Chico'' and
``Centro,'' with SSI-12 being the most relevant variable, then SPI-36 in
``Norte Chico'' and SPI-24 in ``Sur.''
